---
title: "EDA"
format: html
editor: visual
---

## Introduction

The Behavioral Risk Factor Surveillance System (BRFSS) is an annual health-related telephone survey conducted by the CDC, collecting responses from over 400,000 Americans on health behaviors, chronic conditions, and preventative service use since 1984. For this project, trhe dataset diabetes_binary_health_indicators_BRFSS2015.csv contains 253,680 survey responses from the CDC's BRFSS 2015. The target variable, Diabetes_binary, has two classes: 0: No diabetes and 1: Prediabetes or diabetes. The dataset has 22 featured variables. In my analysis, I will look into all of the variables except for Diabetes_binary. The variables that will be used in the EDA are the following; HighBP, HighChol, CholCheck, BMI, Smoker, Stroke, HeartDiseaseorAttack, PhysActivity, Fruits, Veggies, HvyAlcoholConsump, AnyHealthcare, NoDocbcCost, GenHlth, MentHlth, PhysHlth, DiffWalk, Sex, Age, Education, and Income.

The primary purpose of EDA is to understand the data and its structure, identify patterns, spot anomalies, test hypotheses, and check assumptions. This helps in selecting the right modeling techniques and ensures that the data is clean and suitable for analysis. For this analysis, the ultimate goal of modeling is to build a robust and accurate predictive model that can classify individuals into the two categories of the Diabetes_binary variable (0: No diabetes, 1: Prediabetes or diabetes). This model can then be used to identify individuals at risk and potentially guide public health interventions and strategies.

## Data

Use a relative path to import the data. You likely want to convert a lot of the variables to factors with meaningful level names. Check on missingness, etc

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(readxl)
```

```{r}
# Import the data
diabetes_data <- read_excel("~/Downloads/diabetes_binary_health_indicators_BRFSS2015.xlsm")

# Convert variables to factors
diabetes_data$Diabetes_binary <- factor(diabetes_data$Diabetes_binary, levels = c(0, 1), labels = c("No Diabetes", "Prediabetes or Diabetes"))
diabetes_data$HighBP <- factor(diabetes_data$HighBP, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$HighChol <- factor(diabetes_data$HighChol, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$CholCheck <- factor(diabetes_data$CholCheck, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Smoker <- factor(diabetes_data$Smoker, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Stroke <- factor(diabetes_data$Stroke, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$HeartDiseaseorAttack <- factor(diabetes_data$HeartDiseaseorAttack, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$PhysActivity <- factor(diabetes_data$PhysActivity, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Fruits <- factor(diabetes_data$Fruits, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Veggies <- factor(diabetes_data$Veggies, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$HvyAlcoholConsump <- factor(diabetes_data$HvyAlcoholConsump, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$AnyHealthcare <- factor(diabetes_data$AnyHealthcare, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$NoDocbcCost <- factor(diabetes_data$NoDocbcCost, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$DiffWalk <- factor(diabetes_data$DiffWalk, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Sex <- factor(diabetes_data$Sex, levels = c(0, 1), labels = c("Male", "Female"))
diabetes_data$GenHlth <- factor(diabetes_data$GenHlth, levels = 1:5, labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"))

# Convert Age, Education, and Income to factors with meaningful level names
diabetes_data$Age <- factor(diabetes_data$Age, levels = 1:13, labels = c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80 or older"))
diabetes_data$Education <- factor(diabetes_data$Education, levels = 1:6, labels = c("Never attended school", "Elementary", "Some high school", "High school graduate", "Some college", "College graduate"))
diabetes_data$Income <- factor(diabetes_data$Income, levels = 1:8, labels = c("Less than $10,000", "$10,000-$14,999", "$15,000-$19,999", "$20,000-$24,999", "$25,000-$34,999", "$35,000-$49,999", "$50,000-$74,999", "$75,000 or more"))

# Check for missing values
missing_values <- sapply(diabetes_data, function(x) sum(is.na(x)))

# Summary of the dataset
summary(diabetes_data)
```

## Summarizations

You should then produce meaningful summary statistics and plots about the data you are working with (especially as it relates to your response).

Although a best practice would be to split the data at hand into a training and testing set first, go ahead and do your EDA on the full data.

Be sure to have a narrative about what you are exploring and what the summaries and graphs you created say about the relationships in your data.

Distribution of Target Variable The distribution plot of the Diabetes_binary variable shows the count of individuals with and without diabetes. This plot indicates the prevalence of diabetes in the dataset.

```{r}
# Bar plot for the target variable
ggplot(diabetes_data, aes(x = Diabetes_binary)) +
  geom_bar(fill = "blue", color = "black") +
  ggtitle("Distribution of Diabetes Status") +
  xlab("Diabetes Status") +
  ylab("Count")
```

Distribution of BMI The histogram of BMI by diabetes status shows how BMI values are distributed among individuals with and without diabetes. Typically, higher BMI values are associated with a greater risk of diabetes, and we expect to see a higher concentration of diabetes cases in the higher BMI ranges.

```{r}
# Histogram of BMI
ggplot(diabetes_data, aes(x = BMI, fill = Diabetes_binary)) +
  geom_histogram(binwidth = 1, position = "dodge", color = "black") +
  ggtitle("Distribution of BMI by Diabetes Status") +
  xlab("BMI") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))
```

Relationship Between Age and Diabetes Status The bar plot of age distribution by diabetes status helps us understand how age influences the likelihood of diabetes. We expect older age groups to have a higher prevalence of diabetes.

```{r}
# Bar plot for Age distribution by Diabetes Status
ggplot(diabetes_data, aes(x = Age, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("Age Distribution by Diabetes Status") +
  xlab("Age Group") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Relationship Between Physical Activity and Diabetes Status The bar plot for physical activity by diabetes status shows the impact of physical activity on diabetes risk. Regular physical activity is expected to be associated with a lower risk of diabetes.

```{r}
# Bar plot for Physical Activity by Diabetes Status
ggplot(diabetes_data, aes(x = PhysActivity, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("Physical Activity by Diabetes Status") +
  xlab("Physical Activity") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))

```

Proportion of Diabetes by Gender and Proportion of Diabetes by General Health. The proportion table shows the distribution of diabetes status by gender. This can help understand if there is a significant difference in diabetes prevalence between males and females.

The proportion table for general health by diabetes status indicates how self-reported health status correlates with diabetes. Poor general health is likely associated with a higher prevalence of diabetes

```{r}
# Proportion of diabetes by gender
prop_table_gender <- prop.table(table(diabetes_data$Sex, diabetes_data$Diabetes_binary), margin = 1)
print(prop_table_gender)

# Proportion of diabetes by general health
prop_table_genhlth <- prop.table(table(diabetes_data$GenHlth, diabetes_data$Diabetes_binary), margin = 1)
print(prop_table_genhlth)

```

Distribution of General Health by Diabetes Status The bar plot for general health by diabetes status visualizes the relationship between self-reported health status and diabetes. We expect to see poorer health statuses associated with a higher prevalence of diabetes.

```{r}
# Bar plot for General Health by Diabetes Status
ggplot(diabetes_data, aes(x = GenHlth, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("General Health by Diabetes Status") +
  xlab("General Health") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))

```

Relationship Between High Blood Pressure and Diabetes Status The bar plot for high blood pressure by diabetes status shows the correlation between hypertension and diabetes. High blood pressure is a known risk factor for diabetes.

```{r}
# Bar plot for High Blood Pressure by Diabetes Status
ggplot(diabetes_data, aes(x = HighBP, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("High Blood Pressure by Diabetes Status") +
  xlab("High Blood Pressure") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))

```

Relationship Between High Cholesterol and Diabetes Status The bar plot for high cholesterol by diabetes status illustrates the relationship between cholesterol levels and diabetes. High cholesterol is often associated with diabetes.

```{r}
# Bar plot for High Cholesterol by Diabetes Status
ggplot(diabetes_data, aes(x = HighChol, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("High Cholesterol by Diabetes Status") +
  xlab("High Cholesterol") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))

```

Relationship Between Smoking and Diabetes Status The bar plot for smoking by diabetes status shows how smoking status relates to diabetes. Smoking is a risk factor for many chronic diseases, including diabetes.

```{r}
# Bar plot for Smoking by Diabetes Status
ggplot(diabetes_data, aes(x = Smoker, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("Smoking by Diabetes Status") +
  xlab("Smoking Status") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))

```

Conclusion By analyzing these summary statistics and visualizations, we gain insights into how various factors like BMI, age, and physical activity are related to diabetes. This understanding helps in building a predictive model to identify individuals at risk of diabetes effectively. Further analysis can include more variables and interaction effects to improve model accuracy.

These additional summary statistics and visualizations provide further insights into how various health-related factors are associated with diabetes. Understanding these relationships is crucial for building a predictive model that can accurately identify individuals at risk of diabetes. By exploring these variables in depth, we can improve the model's accuracy and reliability.

ADD THE LINK HERE TO CONNECT TO MODELLING WHEN DONE AND GIVE MORE INSIGHT TO THE TAKEAWAYS OF THE EDA AND Be sure to have a narrative about what you are exploring and what the summaries and graphs you created say about the relationships in your data!!!
