---
title: "EDA"
format: html
editor: visual
---

## Introduction

The Behavioral Risk Factor Surveillance System (BRFSS) is an annual health-related telephone survey conducted by the CDC, collecting responses from over 400,000 Americans on health behaviors, chronic conditions, and preventative service use since 1984. For this project, trhe dataset diabetes_binary_health_indicators_BRFSS2015.csv contains 253,680 survey responses from the CDC's BRFSS 2015. The target variable, Diabetes_binary, has two classes: 0: No diabetes and 1: Prediabetes or diabetes. The dataset has 22 featured variables. In my analysis, I will look into all of the variables except for Diabetes_binary. The variables that will be used in the EDA are the following; HighBP, HighChol, CholCheck, BMI, Smoker, Stroke, HeartDiseaseorAttack, PhysActivity, Fruits, Veggies, HvyAlcoholConsump, AnyHealthcare, NoDocbcCost, GenHlth, MentHlth, PhysHlth, DiffWalk, Sex, Age, Education, and Income.

The primary purpose of EDA is to understand the data and its structure, identify patterns, spot anomalies, test hypotheses, and check assumptions. This helps in selecting the right modeling techniques and ensures that the data is clean and suitable for analysis. For this analysis, the ultimate goal of modeling is to build a robust and accurate predictive model that can classify individuals into the two categories of the Diabetes_binary variable (0: No diabetes, 1: Prediabetes or diabetes). This model can then be used to identify individuals at risk and potentially guide public health interventions and strategies.

## Part 1 Reading in Data and converting a lot of the variables to factors with meaningful level names as well checking on missingness in data.

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(readxl)
```

```{r}
# Import the data
diabetes_data <- read_excel("diabetes_binary_health_indicators_BRFSS2015.xlsm")

# Convert variables to factors
diabetes_data$Diabetes_binary <- factor(diabetes_data$Diabetes_binary, levels = c(0, 1), labels = c("No Diabetes", "Prediabetes or Diabetes"))
diabetes_data$HighBP <- factor(diabetes_data$HighBP, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$HighChol <- factor(diabetes_data$HighChol, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$CholCheck <- factor(diabetes_data$CholCheck, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Smoker <- factor(diabetes_data$Smoker, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Stroke <- factor(diabetes_data$Stroke, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$HeartDiseaseorAttack <- factor(diabetes_data$HeartDiseaseorAttack, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$PhysActivity <- factor(diabetes_data$PhysActivity, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Fruits <- factor(diabetes_data$Fruits, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Veggies <- factor(diabetes_data$Veggies, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$HvyAlcoholConsump <- factor(diabetes_data$HvyAlcoholConsump, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$AnyHealthcare <- factor(diabetes_data$AnyHealthcare, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$NoDocbcCost <- factor(diabetes_data$NoDocbcCost, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$DiffWalk <- factor(diabetes_data$DiffWalk, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Sex <- factor(diabetes_data$Sex, levels = c(0, 1), labels = c("Male", "Female"))
diabetes_data$GenHlth <- factor(diabetes_data$GenHlth, levels = 1:5, labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"))

# Convert Age, Education, and Income to factors with meaningful level names
diabetes_data$Age <- factor(diabetes_data$Age, levels = 1:13, labels = c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80 or older"))
diabetes_data$Education <- factor(diabetes_data$Education, levels = 1:6, labels = c("Never attended school", "Elementary", "Some high school", "High school graduate", "Some college", "College graduate"))
diabetes_data$Income <- factor(diabetes_data$Income, levels = 1:8, labels = c("Less than $10,000", "$10,000-$14,999", "$15,000-$19,999", "$20,000-$24,999", "$25,000-$34,999", "$35,000-$49,999", "$50,000-$74,999", "$75,000 or more"))

# Check for missing values
missing_values <- sapply(diabetes_data, function(x) sum(is.na(x)))

# Summary of the dataset
summary(diabetes_data)
```

## Summarizations

In exploring the dataset, I've focused on understanding the distribution of diabetes across various demographic and health-related variables. This investigation is particularly meaningful to me, as diabetes runs in my family on my mother's side, and I am keenly interested in comprehending how this condition correlates with different factors.

Here we are looking into the distribution of Target Variable The distribution plot of the Diabetes_binary variable shows the count of individuals with and without diabetes. This plot indicates the prevalence of diabetes in the dataset.

```{r}
# Bar plot for the target variable
ggplot(diabetes_data, aes(x = Diabetes_binary)) +
  geom_bar(fill = "blue", color = "black") +
  ggtitle("Distribution of Diabetes Status") +
  xlab("Diabetes Status") +
  ylab("Count")
```

Here have histogram highlighting the distribution of BMI. The histogram of BMI by diabetes status shows how BMI values are distributed among individuals with and without diabetes. Typically, higher BMI values are associated with a greater risk of diabetes, and we expect to see a higher concentration of diabetes cases in the higher BMI ranges. Here we can see there is a normal distribution of BMI. Overall, this is an interesting variable that I think will prove to be useful for the analysis.

```{r}
# Histogram of BMI
ggplot(diabetes_data, aes(x = BMI, fill = Diabetes_binary)) +
  geom_histogram(binwidth = 1, position = "dodge", color = "black") +
  ggtitle("Distribution of BMI by Diabetes Status") +
  xlab("BMI") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))
```

Here we are looking into the relationship Between age and diabetes status. I noticed that most of the older family members on my mom's side have diabetes so I'm assuming age is a big indicator but I wanted to look into it. The bar plot of age distribution by diabetes status helps us understand how age influences the likelihood of diabetes. We expect older age groups to have a higher prevalence of diabetes.

```{r}
# Bar plot for Age distribution by Diabetes Status
ggplot(diabetes_data, aes(x = Age, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("Age Distribution by Diabetes Status") +
  xlab("Age Group") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Here we look into the relationship between physical activity and diabetes status. The bar plot for physical activity by diabetes status shows the impact of physical activity on diabetes risk. Regular physical activity is shown to be associated with a lower risk of diabetes.

```{r}
# Bar plot for Physical Activity by Diabetes Status
ggplot(diabetes_data, aes(x = PhysActivity, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("Physical Activity by Diabetes Status") +
  xlab("Physical Activity") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))

```

Here we have simple tables with the proportion of diabetes by gender and proportion of diabetes by General Health. The proportion table shows the distribution of diabetes status by gender. This helped me understand if there is a significant difference in diabetes prevalence between males and females.

Also, the proportion table for general health by diabetes status indicates how self-reported health status correlates with diabetes. Poor general health is shown to be associated with a higher prevalence of diabetes.

```{r}
# Proportion of diabetes by gender
prop_table_gender <- prop.table(table(diabetes_data$Sex, diabetes_data$Diabetes_binary), margin = 1)
print(prop_table_gender)

# Proportion of diabetes by general health
prop_table_genhlth <- prop.table(table(diabetes_data$GenHlth, diabetes_data$Diabetes_binary), margin = 1)
print(prop_table_genhlth)

```

Another plot we looked at was the bar plot highlighting the distribution of general health by diabetes status. The bar plot for general health by diabetes status visualizes the relationship between self-reported health status and diabetes. We expect to see poorer health statuses associated with a higher prevalence of diabetes but it looks like people with diabetes from this sample have pretty good health for the most part.

```{r}
# Bar plot for General Health by Diabetes Status
ggplot(diabetes_data, aes(x = GenHlth, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("General Health by Diabetes Status") +
  xlab("General Health") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))

```

Another bar plot we looked was the relationship between high blood pressure and diabetes. The bar plot for high blood pressure by diabetes status shows the correlation between hypertension and diabetes. High blood pressure is a known risk factor for diabetes so the results of the plot make sense.

```{r}
# Bar plot for High Blood Pressure by Diabetes Status
ggplot(diabetes_data, aes(x = HighBP, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("High Blood Pressure by Diabetes Status") +
  xlab("High Blood Pressure") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))

```

Next, we looked at the relationship between high cholesterol and diabetes. This bar plot for high cholesterol by diabetes status illustrates the relationship between cholesterol levels and diabetes. High cholesterol is often associated with diabetes and that is shown in the plot as well.

```{r}
# Bar plot for High Cholesterol by Diabetes Status
ggplot(diabetes_data, aes(x = HighChol, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("High Cholesterol by Diabetes Status") +
  xlab("High Cholesterol") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))

```

Finally, I was interested in observing the relationship between smoking and diabetes. No one smokes in my family but I watched this ad that said that smoking is a risk factor for diabetes. I found this very interesting because I usually associate it with lung cancer or heart diseases but not diabetes. Looking at the bar plot for smoking by diabetes, it shows how smoking status relates to diabetes.

```{r}
# Bar plot for Smoking by Diabetes Status
ggplot(diabetes_data, aes(x = Smoker, fill = Diabetes_binary)) +
  geom_bar(position = "dodge", color = "black") +
  ggtitle("Smoking by Diabetes Status") +
  xlab("Smoking Status") +
  ylab("Count") +
  scale_fill_manual(values = c("blue", "red"))

```

The visualizations created during the EDA, such as histograms and box plots, illustrate the relationships between diabetes and various predictors. These plots provide a visual representation of how the distributions of different variables differ between those with and without diabetes. For example, a box plot showing higher BMI values among diabetics compared to non-diabetics visually underscores the link between obesity and diabetes risk. By analyzing these summary statistics and visualizations, we gain insights into how various factors like BMI, age, and physical activity are related to diabetes. This understanding helps in building a predictive model to identify individuals at risk of diabetes effectively.

Overall, these exploratory analyses not only offer a better understanding of diabetes distribution among various factors but also inform the development of predictive models. By examining these relationships, we can identify key variables to include in our models and better tailor our approaches to predicting diabetes risk. Based off the plots, I believe the best variables to use in the model include but are not limited to the following; HighBP + HighChol, CholCheck, BMI, Smoker, Stroke and HeartDiseaseorAttack.

[Click here for the Modeling Page](docs/Modeling.html)
